---
title: "PS270 Final Project: Constitutional Design and Political Rights"
author: "Lalitha Viswanathan"
description: "An analysis of how constitutional rigidity and age influence democratic outcomes"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(tidyverse)
judicial_data <- read_csv("data/data.csv")
head(judicial_data)
library(forcats) 
library(ggrepel)
```


```{r americas, include=FALSE}
judicial_data_america <- judicial_data %>% 
  filter(c_amer == 1 | n_amer == 1 | s_amer == 1) %>% 
  mutate(am_region = case_when(
    c_amer == 1 ~ "Central America", 
    n_amer == 1 ~ "North America", 
    s_amer == 1 ~ "South America"
  )) %>% 
  select(-europe, -asia, -africa, -meast, -america, -aurasia)

judicial_data_america <- judicial_data_america %>%
  arrange(am_region, countryn) %>%
  mutate(countryn = fct_inorder(countryn)) 

```

## Introduction

This research explores the relationship between the formal structure of a nation's founding documents and the actual protection of political liberties. Using data from the 2004 Tuck School at Dartmouth paper, "Judicial Checks and Balances," I investigate a central question in institutional design: To what extent does the formal design of a constitution determine the quality of political rights?

Understanding this relationship is vital for emerging democracies. If specific design features, such as constitutional rigidity, can successfully "lock in" democratic gains, then institutional engineering is a powerful tool for progress. I initially hypothesized that a positive relationship exists between constitutional rigidity and political rights. Furthermore, I hypothesized that modern constitutions—informed by contemporary human rights standards—would be more successful at protecting these rights than older documents.

The observed pattern in the data that would provide support for my hypothesis would be that when the rigidity of the constitution goes up, the political rights index also goes up, meaning that when a constitution is more strictly enforced, political rights are protected. For my other hypothesis, when the constitution is older, the political rights index would go down, meaning that older countries would be less successful in protecting political rights than more modern countries with newer constitutions. 


## Data Section

The data for this study is cross-sectional, capturing a snapshot of various countries in 1996. The research design focuses on comparing institutions across different American nations at a single point in time to identify correlations between legal rules and political outcomes. This research is from the Tuck School at Dartmouth, a paper entitled, "Judicial Checks and Balances," published in 2004. 


Measurement of Variables:
Outcome Variable: The Political Rights Index (1996), sourced from Freedom House. It is a 1-7 scale measuring fair elections, the role of the opposition, and self-determination.
Explanatory Variable 1: Rigidity of Constitution, a 1-4 scale measuring the difficulty of amendment (Maddex, 1995).
Explanatory Variable 2: Constitution Year, representing the year the current document was ratified.

Each of the descriptions of the methodology that the researchers used are taken from this website: https://doi.org/10.1086/381480 

The following plot summarizes the distribution of the outcome variable, Political Rights, across the American countries in the sample.


## Visualization displaying the Political Rights Index of countries throughout the Americas, separated by region (North, Central, South)

```{r visualization, echo = FALSE}
visualization <- judicial_data_america %>% 
  ggplot(aes(x = rigid, y = pright, color = countryn)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE, color = "black", size = .75) + 
  geom_text_repel(aes(label = countryn), size = 3, show.legend = FALSE, color = "black") + 
  facet_wrap(~ am_region) +
  labs(
    x = "Rigidity of Constitution",
    y = "Index of political rights in 1996", 
    title = "Political Rights Index and Relationship to Rigidity of\nConstitutions in American Countries (1996)",
    subtitle = "Higher ratings for rigidity of consitution indicate its more difficult to change\nHigher ratings of political rights indicate countries come closer to ideals",
    color = "Country Name", 
    caption = "Rigidity Source: Maddex (1995); Political Rights Source: Freedom House (1996)"
  ) +  scale_color_viridis_d(option = "plasma") +
   theme(legend.position = "none")

visualization
```

This visualization shows that in South America, there is a clear positive correlation (upward-sloping line). Countries like Chile and Ecuador, which have more rigid constitutions (closer to 3.0), also enjoy higher levels of political rights. This supports the theory that making a constitution difficult to change can help "lock in" democratic protections and prevent sudden authoritarian overthrows.

In Central America, the relationship is the opposite or inconsistent. Haiti has the highest rigidity score (4.0) in the entire dataset, yet its political rights score is quite low (3.0). Cuba has a lower rigidity score but the lowest possible rights score. Interpretation: In this region, a "rigid" constitution doesn't necessarily protect people. Sometimes, authoritarian regimes create rigid constitutions not to protect rights, but to make it harder for reformers to change the system.

Canada and the USA represent the "ideal" model in this study: high constitutional rigidity coupled with the highest possible political rights. However, Mexico serves as a reminder that rigidity (3.0) does not automatically grant high rights (4.0), as it lagged behind its northern neighbors in 1996.

The data suggests that Constitutional Rigidity is not a "magic bullet" for democracy. In "stable regions" (South America), rigidity appears to act as a shield that protects political rights. In unstable or authoritarian regions (Central America/Caribbean), high rigidity may simply be a rule on paper that fails to protect citizens in practice, or even a tool used by those in power to prevent change.

Key Observation: The most "rigid" country in the set (Haiti) is far less democratic than countries with "easier to change" constitutions like Panama or Argentina. This suggests that the content of the laws and the culture of the country matter more than just how hard the document is to edit.

## Assessing a relationship between the year the Constitution was written, its rigidity, and political rights

Guiding Question: Does the effectiveness of constitutional rigidity in protecting political rights depend on the age of the constitution throughout America?


```{r, include=FALSE}
library(knitr)
```


```{r new model, include=FALSE}
model <- lm(pright ~ rigid + cons_yr, data = judicial_data_america)
model
summary(model)

library(broom)

```

```{r, echo = FALSE}
tidy(model) %>%
  mutate(across(where(is.numeric), ~round(., 2))) %>%
  kable(caption = "Multiple Linear Regression: Predicting Political Rights")
```


Since the `rigid` coefficient is positive, it means more rigidity leads to more rights. Since the `cons_yr` coefficient is negative, it means newer constitutions have fewer rights. 
For every 1-point increase in the Rigidity of a constitution, the Political Rights index increases by 0.1263 points, holding the year constant. It suggests that making a constitution harder to change has a positive (though relatively small) impact on protecting political rights. For every year later a constitution was written, the Political Rights index decreases by 0.0139 points, holding rigidity constant. A constitution written in 1890 is predicted to have about 1.4 points higher political rights than one written in 1990 (100 x 0.0139), even if they have the exact same rigidity score.

This graph shows Rigidity on the X-axis and Political Rights on the Y-axis. To include the Year, there are two colors: old vs. new.


```{r visualization 2, echo = FALSE}
judicial_data_america <- judicial_data_america %>%
  mutate(Age = ifelse(cons_yr <= 1980, "Established (Pre-1980)", "Recent (Post-1980)"))

regression_plot <- ggplot(judicial_data_america, aes(x = rigid, y = pright)) +
  geom_point(aes(color = Age)) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(
    title = "Rights by Constitutional Rigidity and Age",
    x = "Rigidity Score",
    y = "Political Rights Index",
    color = "Constitutional Era"
  ) +
  theme_minimal()
regression_plot
``` 

The Model: `pright ~ rigid + cons_yr`

Rigidity Coefficient (0.1263): This is represented by the blue regression line. It shows a positive relationship between constitutional rigidity and political rights. The coefficient for rigidity, 0.1263, suggests that 'locking in' a constitution provides a boost to political rights, acting as a structural shield.

Year Coefficient (-0.0139): This is represented by the color-coded clusters. Because the coefficient is negative, the 'Established' (older) countries sit higher on the Y-axis than the 'Recent' (newer) countries, even when they share the same rigidity score. The coefficient for the year of enactment, -0.0139, indicates that newer constitutions are associated with lower political rights scores, suggesting that recent constitutional rewrites often occur in unstable environments compared to older, more established legal frameworks.

The visualization shows that while constitutional design (rigidity) provides a slight upward trend in rights, the 'longevity' of the constitution is a major factor. Contrary to my initial hypothesis, the data suggests that older constitutions outperform modern ones in protecting political rights. While constitutional rigidity provides a structural foundation for rights, it is not a substitute for institutional age. In the Americas, the most robust protections for political rights are found where a rigid design is paired with the stability of an established, long-standing constitution.

The model attempted to predict political rights based on constitutional rigidity and the year of the constitution was not statistically significant if alpha = 0.05. While there is a slight suggestion that older constitutions are associated with higher political rights (p = 0.06), constitutional rigidity had no detectable impact on the outcome. Furthermore, the small sample size limits the power of this analysis. 

## Conclusion 

This analysis finds mixed support for my initial hypotheses. While constitutional rigidity shows a slight positive correlation with political rights, the effect is secondary to the age of the document. Contrary to my second hypothesis, modern constitutions are associated with lower political rights scores in this 1996 snapshot. This suggests that "modern" language is less important than the institutional stability that comes with an older document.

Limitations and Threats to Inference:
The first significant limitation is the small sample size. I wanted to look at just the Americas, and therefore the countries are more limited than looking at a larger continent (Africa, Asia) with a larger sample size as well. Another  significant limitation of this study is omitted variable bias. Factors such as GDP per capita, colonial history, or education levels are likely confounders that influence both when a constitution is rewritten and the level of political rights. Furthermore, this study is observational and cross-sectional. Therefore, we cannot interpret these coefficients causally. We cannot say that "waiting 50 years" causes democracy; rather, older constitutions exist in countries that have already achieved a level of stability that allows a constitution to survive. 
