---
title: "PS270 Final Project"
author: "Lalitha Viswanathan"
description: "My final project"
output:    
  distill::distill_article: 
    self_contained: false
---



## Loading in the data

```{r loading}
library(tidyverse)
judicial_data <- read_csv("data/data.csv")
head(judicial_data)

library(forcats) 
library(ggrepel)
```

## Proposal

This research is from the Tuck School at Dartmouth, a paper entitled, "Judicial Checks and Balances," published in 2004. 

Research question I want to know: To what extent does the formal design of a constitution determine the quality of political rights?

Hypothesis: I hypothesize that there is a positive relationship between constitutional rigidity and political rights. Additionally, I hypothesize that modern constitutions would be more successful at protecting political rights with modern language and beliefs in protections for people. 

The explanatory variables are the rigidity of the constitution and the constitution year. For rigidity, this is how it was measured: Measures (on a scale from 1 to 4) how hard it is to change the constitution in a given country. One point each is given if the approval of the majority of the legislature, the chief of state, and a referendum is necessary in order to change the constitution. An additional point is given for each of the following: if a supermajority in the legislature (more than 66 percent of votes) is needed, if both houses of the legislature have to approve, if the legislature has to approve the amendment in two consecutive legislative terms, or if the approval of a majority of the state legislature is required. Source: Maddex (1995)." The other explanatory variable is the year that the constitution was ratified. 

The outcome variable of interest is the political rights index, and this is how it was measured: "Index of political rights in 1996 (on a scale from 1 to 7). Higher ratings indicate countries that come closer to the ideals suggested by the following checklist: (1) there are free and fair elections, (2) those elected rule, (3) there are competitive parties or other competitive political groupings, (4) the opposition has an important role and power, and (5) the entities have self-determination or an extremely high degree of autonomy. Source: Freedom House (1996)." 

Each of the descriptions of the methodology that the researchers used are taken from this website: https://faculty.tuck.dartmouth.edu/images/uploads/faculty/rafael-laporta/jcb_variable_description.pdf

The observed pattern in the data that would provide support for my hypothesis would be that when the rigidity of the constitution goes up, the political rights index also goes up, meaning that when a constitution is more strictly enforced, political rights are protected. For my other hypothesis, when the constitution is older, the political rights index would go down, meaning that older countries would be less successful in protecting political rights than more modern countries with newer constitutions. 

## Filtering the data to certain regions/continents

```{r americas}
judicial_data_america <- judicial_data %>% 
  filter(c_amer == 1 | n_amer == 1 | s_amer == 1) %>% 
  mutate(am_region = case_when(
    c_amer == 1 ~ "Central America", 
    n_amer == 1 ~ "North America", 
    s_amer == 1 ~ "South America"
  )) %>% 
  select(-europe, -asia, -africa, -meast, -america, -aurasia)

judicial_data_america <- judicial_data_america %>%
  arrange(am_region, countryn) %>%
  mutate(countryn = fct_inorder(countryn)) 

```

```{r middle east}
judicial_data_meast <- judicial_data %>% 
  filter(meast == 1) %>% 
  mutate(Region = case_when(
    meast == 1 ~ "Middle East")) %>% 
  select(-europe, -asia, -africa, -america, -aurasia, -oceania, -c_amer, -s_amer, -n_amer)
```


## Visualization displaying the Political Rights Index of countries throughout the Americas, separated by region (North, Central, South)

```{r visualization}
visualization <- judicial_data_america %>% 
  ggplot(aes(x = rigid, y = pright, color = countryn)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE, color = "black", size = .75) + 
  geom_text_repel(aes(label = countryn), size = 3, show.legend = FALSE, color = "black") + 
  facet_wrap(~ am_region) +
  labs(
    x = "Rigidity of Constitution",
    y = "Index of political rights in 1996", 
    title = "Political Rights Index and Relationship to Rigidity of\nConstitutions in American Countries (1996)",
    subtitle = "Higher ratings for rigidity of consitution indicate its more difficult to change\nHigher ratings of political rights indicate countries come closer to ideals",
    color = "Country Name", 
    caption = "Rigidity Source: Maddex (1995); Political Rights Source: Freedom House (1996)"
  ) +  scale_color_viridis_d(option = "plasma") +
   theme(legend.position = "none")

visualization
```

This visualization shows that in South America, there is a clear positive correlation (upward-sloping line). Countries like Chile and Ecuador, which have more rigid constitutions (closer to 3.0), also enjoy higher levels of political rights. This supports the theory that making a constitution difficult to change can help "lock in" democratic protections and prevent sudden authoritarian overthrows.

In Central America, the relationship is the opposite or inconsistent. Haiti has the highest rigidity score (4.0) in the entire dataset, yet its political rights score is quite low (3.0). Cuba has a lower rigidity score but the lowest possible rights score. Interpretation: In this region, a "rigid" constitution doesn't necessarily protect people. Sometimes, authoritarian regimes create rigid constitutions not to protect rights, but to make it harder for reformers to change the system.

Canada and the USA represent the "ideal" model in this study: high constitutional rigidity coupled with the highest possible political rights. However, Mexico serves as a reminder that rigidity (3.0) does not automatically grant high rights (4.0), as it lagged behind its northern neighbors in 1996.

The data suggests that Constitutional Rigidity is not a "magic bullet" for democracy. In "stable regions" (South America), rigidity appears to act as a shield that protects political rights. In unstable or authoritarian regions (Central America/Caribbean), high rigidity may simply be a rule on paper that fails to protect citizens in practice, or even a tool used by those in power to prevent change.

Key Observation: The most "rigid" country in the set (Haiti) is far less democratic than countries with "easier to change" constitutions like Panama or Argentina. This suggests that the content of the laws and the culture of the country matter more than just how hard the document is to edit.

## Assessing a relationship between the year the Constitution was written and the rigidity

Guiding Question: Do older, established constitutions protect rights better, or do modern constitutions perform better? 

```{r}

library(knitr)

table_data <- judicial_data %>%
  mutate(age_group = case_when(
    cons_yr < 1950 ~ "Over 45 Years Old",
    cons_yr >= 1950 & cons_yr < 1985 ~ "10–45 Years Old",
    cons_yr >= 1985 ~ "Recent (Under 10 Years)"
  )) %>%
  group_by(age_group) %>%
  summarise(
    `Number of Countries` = n(),
    `Average Political Rights` = mean(pright, na.rm = TRUE),
    `Average Rigidity` = mean(rigid, na.rm = TRUE)
  ) %>%
  arrange(desc(`Average Political Rights`))

table <- table_data %>% 
  drop_na(age_group) %>% 
  select(age_group, `Average Political Rights`) %>% 
  pivot_wider(
    names_from = age_group,
    values_from = `Average Political Rights`
  )

knitr::kable(table, digits = 2, caption = "Average Political Rights Score for Constitutions Based on Age Group")

```

These numbers show that Constitutional Age is a stronger shield for democracy than Constitutional Design. Countries with 'legacy' constitutions (pre-1950) maintain a stronger political rights average of 6.47. Conversely, there is almost no difference in rights between constitutions written in the 1960s (3.94) and the 1990s (3.90). 

New countries are writing the rules of democracy, but the strength of the institution is not nearly as much as established, democratic powers. The newness of a country's law is a sign of insecurity.
